\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-17 chapters/wcble.tex
Proofread mathematical notation
Indicated comments Reinhold and Fabien

\chapter{\Acs{erp} latency estimation and alignment}
\label{sec:wcble}
\emph{\Cref{sec:wcble/literature/contrib}, \cref{sec:wcble/methods} (including
\cref{fig:diagram-train,fig:diagram-eval}), and the third-to-last paragraph of
\cref{sec:wcble/conclusion} were published
as part of~\textcite{VanDenKerchove2024}.}

@@

\Ac{eeg}-based visual oddball \acp{bci} establish communication with paralyzed
individuals by decoding \acp{erp} obtained from time-modulated flashing
stimuli (see~\cref{fig:bci/loop}).
These \acp{erp} consist of multiple \ac{erp}-components, peaks and troughs in
\ac{erp} waveform (see \cref{fig:bci/erp}).
These components each have their own distinct amplitude and latency.
\Ac{erp} component latency can reflect a wide range of cognitive,
neurophysiological, and clinical properties, that can be confounding factors in
the development of \acp{bci}~\cite{Luck2014}.
Latency is often assessed as the timing of an \ac{erp} component peak, obtained
after averaging over trials.
Latencies of non-averaged, single-trial \acp{erp} should capture even more
relevant information, yet they are very hard to measure due to the low \ac{snr} of
single-trial \acp{erp}.
Within-session variability of \ac{erp} latency, called \ac{erp} latency
\emph{jitter}, has been shown to influence \ac{bci}
accuracy~\cite{Thompson2012}.
This effect is especially prominent when bringing \ac{bci} development from the
lab to the clinical setting.
For instance, \textcite{Zisk2021} showed that \ac{als} results significantly
increased \ac{erp} jitter.
\textcite{Arico2014} showed that latency jitter negatively affects \ac{bci}
performance, and disproportionately so when the user is not directly gazing at
intended targets, such as is the case for individuals with \ac{sspgi}.
Therefore, properly accounting for \ac{erp} latency jitter can contribute to
the development of gaze-independent visual \acp{bci}.

@@

The source time course is defined by the following function:
\begin{equation}
  s(t,l) =
  \begin{cases}
    a\sin\left(2\pi f\left(t-l\right)\right) & \text{if  $l-\frac{1}{2f} < t < l-\frac{1}{2f}$} \\
    0 & \text{otherwise}
  \end{cases}
\end{equation}
over time $t$, with latency $l$, amplitude $a=\num{1e{-7}}$V, and frequency
$f=4$.

% ERROR BETWEEN HERE ==========================================================
100 target epochs containing the evoked potential and noise and 100
non-target epochs containing only noise were simulated.
The target epochs were jittered by setting the latency offset $l$
of each target epoch to a random latency drawn from a normal distribution with 0 mean and standard
deviations of respectively $\sigma=0.1,0.2,0.3$s.
The noise was scaled at 32 different noise levels, with the \ac{snr} of a
target epoch ranging from 0 to -31dB, with
\begin{equation}
  \text{SNR} = 10\log_{10}\frac{\text{var}s(t,0)}{\text{var}\left(\text{noise}\right)}
\end{equation}

@@

\Cref{fig:wcble/results/accuracy} shows that the classification accuracy of both \ac{tlda} and \ac{cble}
decreases with
$\sigma$ and \ac{snr}, but \ac{cble} outperforms \ac{tlda}.
This difference is particularly large when $\sigma$ is higher ($\sigma=0.2$
and $\sigma=0.3$) and \ac{snr} is not yet high.
\Ac{wcble} always scored better than \ac{cble} and \ac{tlda}, except for
$\sigma=0.1$ and $\text{\ac{snr}} < 27 \text{dB}$, at which point both \ac{cble}
and \ac{wcble} performed at chance level.
When considering the accuracy of \ac{wcble}, we see that it shows more of a
plateau as \ac{snr} increases.

@@

The key finding of the simulation study presented
in \cref{sec:wcble/results} is that \ac{wcble} is robust to higher evoked
potential jitter and lower \ac{snr} than \ac{cble}.
It could therefore be a better candidate to apply in \ac{erp} analysis and
decoding in those cases where jittered \ac{erp} data is of interest.
At a certain noise threshold, both methods begin to fail, but \ac{wcble}
delivers higher accuracy at intermediate \ac{snr} levels than \ac{cble}, likely due to the iterative alignment process.

>>>>

\chapter{\Acs{erp} latency estimation and alignment}
\label{sec:wcble}
\epigraph{``The sands of time cannot be stopped''}{Brom in Eragon --
Christopher Paolini}
\noindent\emph{\Cref{sec:wcble/literature/contrib}, \cref{sec:wcble/methods} (including
\cref{fig:diagram-train,fig:diagram-eval}), and the third-to-last paragraph of
\cref{sec:wcble/conclusion} were published
as part of~\textcite{VanDenKerchove2024}.}

@@

\Ac{eeg}-based visual oddball \acp{bci} establish communication with paralyzed
individuals by decoding \acp{erp} obtained from time-modulated flashing
stimuli (see~\cref{fig:bci/loop}).
These \acp{erp} consist of multiple \ac{erp}-components, peaks and troughs in
\ac{erp} waveform (see \cref{fig:bci/erp}).
These components each have their own distinct amplitude and latency.
\Ac{erp} component latency can reflect a wide range of cognitive,
neurophysiological, and clinical properties, that can be confounding factors in
the development of \acp{bci}~\cite{Luck2014}.
Latency is often assessed as the timing of an \ac{erp} component peak, obtained
after averaging over trials.
Latencies of non-averaged, single-trial \acp{erp} should capture even more
relevant information, yet they are very hard to measure due to the low \ac{snr} of
single-trial \acp{erp}.
Within-session variability of \ac{erp} latency, called \ac{erp} latency
\emph{jitter}, has been shown to influence \ac{bci}
accuracy~\cite{Thompson2012}.
This effect is especially prominent when bringing \ac{bci} development from the
lab to the clinical setting.
For instance, \textcite{Zisk2021} showed that \ac{als} results in significantly
increased \ac{erp} jitter\todo{qualitative}.
\textcite{Arico2014} showed that latency jitter negatively affects \ac{bci}
performance, and disproportionately so when the user is not directly gazing at
intended targets, such as is the case for individuals with \ac{sspgi}.
Therefore, properly accounting for \ac{erp} latency jitter can contribute to
the development of gaze-independent visual \acp{bci}.

@@

The source time course is defined by the following function:
\begin{equation}
  s(t,l) =
  \begin{cases}
    a\sin\left(2\pi f\left(t-l\right)\right) & \text{if  $l-\frac{1}{2f} < t < l+\frac{1}{2f}$} \\
    0 & \text{otherwise}
  \end{cases}
\end{equation}
over time $t$, with latency $l$, amplitude $a=\num{1e{-7}}$V, and frequency
$f=4$.

% ERROR BETWEEN HERE ==========================================================
100 target epochs containing the evoked potential and noise and 100
non-target epochs containing only noise were simulated.
The target epochs were jittered by setting the latency offset $l$
of each target epoch to a random latency drawn from a normal distribution with 0 mean and standard
deviations of respectively $\sigma=0.1,0.2,0.3$s.
The noise was scaled at 32 different noise levels, with the \ac{snr} of a
target epoch ranging from 0 to -31dB, with
\begin{equation}
  \text{SNR} = 10\log_{10}\frac{\var\left[s(t,0)\right]}{\var\left(\text{noise}\right)}
\end{equation}

@@

\Cref{fig:wcble/results/accuracy} shows that the classification accuracy of both \ac{tlda} and \ac{cble}
decreases with
$\sigma$ and \ac{snr}, but \ac{cble} outperforms \ac{tlda}.
This difference is particularly large\todo{qualitative} when $\sigma$ is higher ($\sigma=0.2$
and $\sigma=0.3$) and \ac{snr} is not yet high.
\Ac{wcble} always scored better than \ac{cble} and \ac{tlda}, except for
$\sigma=0.1$ and $\text{\ac{snr}} < 27 \text{dB}$, at which point both \ac{cble}
and \ac{wcble} performed at chance leveltodo{qualitative}.
When considering the accuracy of \ac{wcble}, we see that it shows more of a
plateau as \ac{snr} increases\todo{qualitative}.

@@

The key finding of the simulation study presented
in \cref{sec:wcble/results} is that \ac{wcble} is robust to higher evoked
potential jitter and lower \ac{snr} than \ac{cble}\todo{qualitative}.
It could therefore be a better candidate to apply in \ac{erp} analysis and
decoding in those cases where jittered \ac{erp} data is of interest.
At a certain noise threshold\todo{qualitative}, both methods begin to fail, but \ac{wcble}
delivers higher accuracy at intermediate \ac{snr} levels than
\ac{cble}\todo{qualitative}, likely due to the iterative alignment process.

### 2024-11-17 figures/gaze_independence/lis_venn.tikz.tex
Update bttda formulation
Indicated comments Reinhold and Fabien

    % Draw the outer lightgray box with ultra thick darkgray line
    \draw[fill=lightgray, draw=darkgray, ultra thick] (-4,-1.5) rectangle (4,5.5);
    % Label for the outer box in the top left corner
    \node[anchor=north west, text=darkgray] at (-4,5.5) {\textbf{\acs{bci} users}};

    % Draw the outermost circle for SSPI with ultra thick lines and bold text matching its line color
    \draw[ultra thick, accent1] (0,2) circle (3);
    \node[text=accent1] at (0,4) {\textbf{\acs{sspi}}};  % Centered bold text with line color

    % Draw the middle circle for SSPGI with ultra thick lines and bold text matching its line color
    \draw[ultra thick, accent2] (0,1) circle (2);
    \node[text=accent2] at (0,1.66) {\textbf{\acs{sspgi}}};  % Centered bold text with line color

    % Draw the innermost circle for LIS with ultra thick lines and bold text matching its line color
    \draw[ultra thick, accent3] (0,0) circle (1);
    \node[text=accent3] at (0,0) {\textbf{\acs{lis}}};  % Centered bold text with line color

>>>>

    % Draw the outer lightgray box with ultra thick darkgray line
    \draw[draw=darkgray, ultra thick] (-4,-1.5) rectangle (4,5.5);
    % Label for the outer box in the top left corner
    \node[anchor=north west, text=darkgray] at (-4,5.5) {\textbf{\acs{bci} users}};

    % Draw the outermost circle for SSPI with ultra thick lines and bold text matching its line color
    \draw[ultra thick, accent1] (0,2) circle (3);
    \node[text=accent1, text width=15em, anchor=center, align=center] at (0,4)
    {\small\textbf{severe speech and \\ physical impairment (SSPI)}};  % Centered bold text with line color

    % Draw the middle circle for SSPGI with ultra thick lines and bold text matching its line color
    \draw[ultra thick, accent2] (0,1) circle (2);
    \node[text=accent2, text width=10em, anchor=center, align=center] at
    (0,2) {\small\textbf{severe speech, \\ physical and gaze \\ impairment
    (SSPGI)}};  % Centered bold text with line color

    % Draw the innermost circle for LIS with ultra thick lines and bold text matching its line color
    \draw[ultra thick, accent3] (0,0) circle (1);
    \node[text=accent3, text width=5em, anchor=center, align=center] at (0,-.1)
    {\small\textbf{Locked-in \\ Syndrome \\ (LiS)}};  % Centered bold text with line color
